\section{Introduction}
\label{sec:introduction}

To audit any computational process, we need robust and well-founded notions of \emph{provenance} to track how data are used. This allows us to answer questions like ``Where did these data come from?'', ``Why are these data in the output?'' and ``How were these data computed?''. Provenance tracking has a wide range of applications, from debugging and program comprehension~\cite{buneman95,cheney07} to improving reproducibility and transparency in scientific workflows~\cite{kontogiannis08}. \emph{Program slicing}, first proposed by~\citet{weiser81}, is a collection of techniques for provenance tracking that attempts to take a run of a program and areas of interest in the output, and turn them into the subset of the input and the program that were responsible for generating those specific outputs.

Existing approaches to program slicing are often tied to particular programming languages or implementations.
In this paper we develop a general categorical approach to program slicing, focusing on a particular technique
called \GPS, where the set of slices of a given value form a lattice of approximations and the forward and
backward slicing procedures generate Galois connections between these lattices. Our main contribution is that
this approach can be seen as a generalised form of automatic differentiation, with slices of values playing
the role of tangents. Our categorical approach should provide a suitable setting for enabling ``automatic''
data provenance for a variety of programming languages, and is easily configured to use alternative
approximation strategies, including quantitative forms of slicing.

\subsection{Galois Program Slicing}
\label{sec:introduction:galois-slicing}

Perera and collaborators introduced the idea of {\em Galois program slicing} as a particular conception of program slicing for provenance, described in several publications~\cite{perera12a,perera16d,ricciotti17}. Galois program slicing (hereafter simply {\emph \GPS}) forms the basis of the open source data visualisation tool \href{https://f.luid.org/}{Fluid}~\cite{perera2025fluid} that allows interactive exploration of programmatically generated visualisations.

At a high level, \GPS assumes that, for each possible value that may be input or output by a program, there exists a lattice of {\em approximations} of that value. For a particular run of a program that takes input $x$ and produces output $y$, we also get a Galois connection between the lattice of approximations of $x$ and the lattice of approximations of $y$. The right half of the Galois connection is the ``forward direction'' taking approximations of the input to approximations of the output; the left half of the Galois connection is the ``backward direction'' that takes approximations of the output to the least (i.e., most approximate) approximation of the input that gives rise to this output approximation. This becomes {\em program slicing} by including the source code of the program as part of the input; then, in the backward direction, the least approximation of the input required for an output approximation includes the least part of the program required as well.

\begin{example}
  \label{ex:introduction-example}
  The following program is written in Haskell syntax \cite{haskell}, using a list comprehension to filter a list of pairs of labels and numbers to those numbers with a given label, and then computing the sum of the numbers:
  \begin{displaymath}
    \begin{array}{l}
      \mathrm{query} :: \mathrm{Label} \to [(\mathrm{Label}, \mathrm{Int})] \to \mathrm{Int} \\
      \mathrm{query}\,l\,\mathit{db} = \mathrm{sum}\,[ n \mid (l',n) \leftarrow \mathit{db}, l \equiv l' ]
    \end{array}
  \end{displaymath}
  With $\mathit{db} = [(\mathsf{a}, 0), (\mathsf{b}, 1), (\mathsf{a}, 1)]$, we will have $\mathit{query}\,\mathsf{a}\,\mathit{db}$ and $\mathit{query}\,\mathsf{b}\,\mathit{db}$ both evaluating to $1$.

  Now suppose that for a given run of the program, we are interested in which of the numerical parts of the input are used to compute the output for the query parameters $l = \mathsf{a}$ and $l = \mathsf{b}$. We can use \GPS to do this. We arrange for the approximations of the input to form the following lattice, where the actual piece of data is at the top and information lost by approximation is represented by $\bot$s:
  \begin{center}
    \begin{tikzpicture}[node distance=0.75cm]
      \node (top) at (0,0) {$[(\mathsf{a}, 0), (\mathsf{b}, 1), (\mathsf{a}, 1)]$};
      % row 2
      \node [below of=top] (ioi) {$[(\mathsf{a}, 0), (\mathsf{b}, \bot), (\mathsf{a}, 1)]$};
      \node [left of=ioi,xshift=-3cm] (oii) {$[(\mathsf{a}, \bot), (\mathsf{b}, 1), (\mathsf{a}, 1)]$};
      \node [right of=ioi,xshift=3cm] (iio) {$[(\mathsf{a}, 0), (\mathsf{b}, 1), (\mathsf{a}, \bot)]$};
      % row 3
      \node [below of=ioi] (oio) {$[(\mathsf{a}, \bot), (\mathsf{b}, 1), (\mathsf{a}, \bot)]$};
      \node [left of=oio,xshift=-3cm] (ooi) {$[(\mathsf{a}, \bot), (\mathsf{b}, \bot), (\mathsf{a}, 1)]$};
      \node [right of=oio,xshift=3cm] (ioo) {$[(\mathsf{a}, 0), (\mathsf{b}, \bot), (\mathsf{a}, \bot)]$};
      % row 4
      \node [below of=oio] (bot) {$[(\mathsf{a}, \bot), (\mathsf{b}, \bot), (\mathsf{a}, \bot)]$};

      % links
      \draw (top) -- (ioi);
      \draw (top) -- (oii);
      \draw (top) -- (iio);
      \draw (ioi) -- (ooi);
      \draw (ioi) -- (ioo);
      \draw (oii) -- (ooi);
      \draw (oii) -- (oio);
      \draw (iio) -- (ioo);
      \draw (iio) -- (oio);
      \draw (ooi) -- (bot);
      \draw (oio) -- (bot);
      \draw (ioo) -- (bot);
    \end{tikzpicture}
  \end{center}
  In both runs of the program, the output approximation lattice looks like this, where $1$ is the actual data point that was returned, and $\bot$ indicates that we are approximating this piece of data away:
  \begin{center}
    \begin{tikzpicture}[node distance=0.75cm]
      \node (top) at (0,0) {$1$};
      \node [below of=top] (bot) {$\bot$};
      \draw (top) -- (bot);
    \end{tikzpicture}
  \end{center}
  These are not the only choices of approximation lattices that we could have made. For the input, we have chosen a lattice that allows us to ``forget'' (approximate away) numbers in the input, but not the labels or the structure of the list itself. However, other choices are also useful. Indeed, one of the aims of this work is to clarify how to choose an approximation structure appropriate for different tasks by use of type information. We elaborate on this further in \secref{semantic-gps}.

  \GPS associates with each run of the program a Galois connection telling us how the inputs and outputs are related in that run. The backwards portion $\partial (\mathit{query}\,l)_r$ tells us, given an approximation of the output, what the least approximation of the input is needed to generate that output. In the case of the two runs considered in this example, if we say we are not interested in the output by feeding in the least approximation $\bot$, then we find that we only need the least approximation of the input:
  \begin{displaymath}
    \partial (\mathit{query}\,l\,\mathit{db})_r(\bot) = [(\mathsf{a},\bot), (\mathsf{b}, \bot), (\mathsf{a}, \bot)]
  \end{displaymath}
  for both $l = \mathsf{a}$ and $l = \mathsf{b}$. If instead we take the greatest approximation of the output (i.e., the output ``$1$'' itself), then the two query runs' backwards approximations return different results:
  \begin{displaymath}
    \begin{array}{l}
      \partial (\mathit{query}\,\mathsf{a}\,\mathit{db})_r(1) = [(\mathsf{a},0), (\mathsf{b},\bot), (\mathsf{a},1)] \\
      \partial (\mathit{query}\,\mathsf{b}\,\mathit{db})_r(1) = [(\mathsf{a},\bot), (\mathsf{b},1), (\mathsf{a},\bot)]
    \end{array}
  \end{displaymath}
  Pieces of the input that were {\em not} used are replaced by $\bot$. As we expect, the run of the query with label $\mathsf{a}$ depends on the entries in the database labelled with $\mathsf{a}$, and likewise for the run with label $\mathsf{b}$.

  In this case, the forwards portion of the Galois connection tells us, for each approximation of the input, whether or not it is sufficient to compute the output. If we provide insufficient data to compute the output, then we will get an underapproximated output. Here for example we will find that $\partial (\mathit{query}\, \mathsf{a})_f([(\mathsf{a},0),(\mathsf{b},\bot),(\mathsf{a},\bot)]) = \bot$ because we need all the values associated with the label $\mathsf{a}$ to compute their sum.

  In a simple query like this, it is easy to work out the dependency relationship between the input and output. However, the benefit of \GPS, and other language-based approaches, is that it is {\em automatic} for all programs, no matter how complex the relationship between input and output. Moreover, by changing what we mean by ``approximation'' we can compute a range of different information about a program.
\end{example}

\subsection{Galois Slicing and Automatic Differentiation}

Previous work on \GPS used a special operational semantics to generate a trace of each execution, and then uses that trace to compute the Galois connections described above, by re-running forwards or backwards over the trace. It would be useful to have a denotational account of \GPS as well, especially if we could provide a semantics where the backwards analysis is baked in, rather than provided by a separately defined ``backwards evaluation'' operation. Our thesis, developed in \secref{approx-as-tangents} and \secref{models-of-total-gps} is that there is a close analogy between \GPS and \emph{automatic differentiation} for differentiable programs~\cite{siskind08,elliott18,vakar22}, which points to a way to develop such an approach. We have already hinted at this in the description above, but let us now make it explicit.

\begin{itemize}[leftmargin=\enummargin]
\item For \GPS, we assume that every value has an associated lattice of {\em approximations}. For differentiable programs, every point has an associated vector space of {\em tangents}.
\item For \GPS, every program has an associated forward approximation map that takes approximations forward from the input to the output. This map {\em preserves meets}. For differentiable programs, every program has a forward derivative that takes tangents of the input to tangents of the output. The forward derivative map is {\em linear}, so it preserves addition of tangents and the zero tangent.
\item For \GPS, every program has an associated backward approximation map that takes approximations of the output back to least approximations of the input. This map {\em preserves joins}. For differentiable programs, every program has a reverse derivative that takes tangents of the output to tangents of the input. This map is again {\em linear}.
\item For \GPS, the forward and backward approximation maps are related by being a Galois connection. For differentiable programming, the forward and reverse derivatives are related by being each others' transpose.
\end{itemize}

Given this close connection between \GPS and differentiable programming, we can take structures intended for modelling automatic differentiation, such as V{\'a}k{\'a}r's CHAD framework and use them to model \GPS. This will enable us to generalise and expand the scope of \GPS to act as a foundation for data provenance in a wider range of computational settings.

\subsection{Outline and Contributions}

%Our primary contribution in this article is to elucidate \GPS by relating it to differentiable programming and automatic differentiation. In doing so, we aim to expand the applicability of program slicing, and to potentially transfer efficient automatic differentiation implementation techniques from automatic differentiation to program slicing and provenance tracking.

\GPS, as any program slicing technique, essentially rests on an analysis of how programs intensionally explore their input, in addition to their extensional behaviour. Such analysis has been carried out over many years in Domain Theory. In \secref{approx-as-tangents}, we use ideas from \citet{berry79}'s \emph{stable domain theory} and develop an analogy between stable functions and smooth functions from mathematical analysis, where stable functions provide a kind of semantic provenance analysis. In \secref{models-of-total-gps}, we abstract from stable functions using V{\'a}k{\'a}r \etal{}'s CHAD framework \cite{vakar22,nunes2023} to build models of a higher-order language that automatically compute slices. We apply this to a concrete higher-order language in \secref{language} and demonstrate the use of the model on variations of \exref{introduction-example}, highlighting the flexibility of our approach. In particular, we show how type structure can be used to control the approximation lattices associated with data points, something that was ``hard coded'' in previous presentations of \GPS. We prove two correctness properties in \secref{definability}, relating the higher-order interpretations to first-order ones, proving the crucial Galois connection property. \secref{related-work} and \secref{conclusion} discuss additional related and future work.

% \begin{enumerate}[leftmargin=\enummargin]
% \item We explain how the CHAD framework of V치k치r et al. can be adapted to give a general categorical framework for \GPS.
% \item Using our adapted CHAD framework, we can explain how type structure can be used to control the approximation lattices associated with data points, something that was ``hard coded'' in previous presentations of \GPS.
% \item With the benefit of our abstract setting, we can relate \GPS to other parts of denotational semantics. In particular, we show that there is a close connection with Stable Domain Theory, a proposed solution to capturing sequentiality by recording more intensional information about programs' sensitivity to approximations.
% \end{enumerate}

We have formalised our major results in Agda, resulting in an executable implementation built directly from the categorical constructions that we have used to compute the examples in \secref{language:examples}. Please consult the file \texttt{everything.agda} in the supplementary material.


% \begin{enumerate}
% \item We explain how the CHAD framework of V치k치r et al. can be adapted to give a general categorical framework for \GPS.
% \item Using our adapted CHAD framework, we can explain how type structure can be used to control the approximation lattices associated with data points, something that was ``hard coded'' in previous presentations of \GPS.
% \item With the benefit of our abstract setting, we can relate \GPS to other parts of denotational semantics. In particular, we show that there is a close connection with Stable Domain Theory, a proposed solution to capturing sequentiality by recording more intensional information about programs' sensitivity to approximations.

% We have \anonymise{\href{https://github.com/bobatkey/approx-diff}{formalised this work in Agda}}{formalised this work in Agda}. Not only does this mean that the constructions and proofs have been checked, but also the construction is executable and can be run on small examples. We are also planning to use Agda's JavaScript backend to produce a demo.
