\section{Higher-Order Language}
\label{sec:language}

To model Galois slicing semantically for higher-order programs, we define a simple total functional
programming language, extending the simply-typed lambda calculus. The language is parameterised by a signature
$\Sigma = (\PrimTy, \PrimOp)$ consisting of a set $\PrimTy$ of base types $\rho$ and a family of sets
$\PrimOp^\rho_{\rho_1,\ldots,\rho_n}$ of primitive operations $\phi$ of arity $n$ over those base types.

\subsection{Syntax}
\label{sec:language:syntax}

The syntax is defined in \figref{syntax}. Types includes base types $\rho$ drawn from $\PrimTy$, along with
standard type formers for sums, products, functions and lists% , plus a type constructor $\tyLift$ to allow
% approximation points to be added explicitly, as discussed in \secref{models-of-total-gps}
. Terms include
variables, the usual introduction and elimination forms% , a monadic return and bind for lifted terms
, and
primitive operations $\phi$.

The language is intentionally minimal: it excludes general recursion, and general inductive or coinductive
types, which we will consider in future work (\secref{conclusion}). Typing judgments for terms are standard
and shown in \figref{typing}, with the usual rules for products, sums, functions, and lists.

\input{fig/syntax}
\input{fig/typing}

\subsection{Semantics}
\label{sec:language:semantics}

\input{fig/semantics}

An interpretation of a signature $\Sigma = (\PrimTy, \PrimOp)$ can be given in any category $\cat{C}$ with
finite products, and assigns to each base type $\rho \in \PrimTy$ an object $\sem{\rho}_{\PrimTy}$ in
$\cat{C}$, and to each primitive operation $\phi \in \PrimOp^\rho_{\rho_1,\ldots,\rho_n}$, a morphism
$\sem{\phi}_{\Op}: \sem{\rho_1}_{\PrimTy} \times \ldots \times \sem{\rho_n}_{\PrimTy} \to
\sem{\rho}_{\PrimTy}$.

Assuming that $\cat{C}$ is bicartesian closed and has a list object (\eqnref{lists}), then we can extend an
interpretation of a signature $\Sigma$ to an interpretation of the whole language over
$\Sigma$. \figref{semantics:types} and \figref{semantics:contexts} define the interpretation of types and
contexts as objects of $\cat{C}$ respectively. Terms are interpreted as morphisms between the interpretations
of the context and type, as defined in \figref{semantics:terms}. We have used the notations $\pi_i$ for
projections, $\prodM{f}{g}$ for pairing, $\coprodM{f}{g}$ for (parameterised) copairing, $!_x$ for morphisms
to the terminal object, and $\lambda$ and $\eval$ for currying and evaluation for exponentials.

For the first-order definability result in \secref{definability}, we will need another interpretation
$\sem{-}_{\mathit{fo}}$ of the first-order types (those constructed from primitive types, sums, unit and
products) in any bicartesian category. Such interpretations are preserved by finite coproduct and coproduct
preserving functors:

\begin{lemma}\label{lem:first-order-agreement-types}
  If $\cat{C}$ and $\cat{D}$ are bicartesian and bicartesian closed categories with interpretations of the
  signature $\Sigma$, $F : \cat{C} \to \cat{D}$ is a bicartesian functor, and
  $F(\sem{\rho}_{\PrimTy}) \cong \sem{\rho}_{\PrimTy}$ for all $\rho$, then for all first-order types $\tau$,
  $F(\sem{\tau}_{\mathit{fo}}) \cong \sem{\tau}$, and similar for contexts only containing first-order types.
\end{lemma}

\subsubsection{Interpretation for Higher-Order \GPS}
\label{sec:language:gps-interpretation}

Given the above, we can now interpret the language in any of the bicartesian closed categories with list
objects we constructed in \secref{models-of-total-gps}. Specifically, we assume that we have an interpretation
of our chosen signature in $\Fam(\LatGal)$. Signatures are first-order, so it does not matter that
$\Fam(\LatGal)$ is not closed. Any such interpretation can be transported to
$\Fam(\MeetSLat \times \JoinSLat^\op)$ along the functor $H$ from \propref{ho-embedding} because it preserves
finite products. We can then interpret the whole language in $\Fam(\MeetSLat \times \JoinSLat^\op)$.

Interpreting a whole program $\Gamma \vdash t : \tau$ yield morphisms in
$\Fam(\MeetSLat \times \JoinSLat^\op)$ which, as in \secref{semantic-gps}, are triples
$(f, \partial f_f, \partial f_r)$ of the underlying function and the forward and backward approximation
maps. However, unlike in $\Fam(\LatGal)$, it is not guaranteed that the forward and backward maps even operate on
the same lattices, let alone form a Galois connection. \lemref{first-order-agreement-types} guarantees that
the lattices agree, but the fact that the pair form a Galois connection is less trivial. We will prove this
property in \secref{definability}.

% Given a model of $\Sigma$ in $\Fam(\LatGal)$, the category $\Fam(\MeetSLat \times \JoinSLat^\op)$ can
% implement $\Sem$. It is bicartesian closed by \corref{mslat-jslat-bcc}, and can model $\tyList$ using infinite
% coproducts (\secref{models-of-total-gps:coproducts-and-products}). The embedding $\HoEmbed: \Fam(\LatGal) \to
% \Fam(\MeetSLat \times \JoinSLat^\op)$ from \propref{ho-embedding} preserves finite products, and so can
% transport each $\sem{\rho}_{\PrimTy}$ and $\sem{\phi}_{\Op}$ into the higher-order setting to provide a model
% of $\Sigma$. Finally the lifting monad in $\Fam(\LatGal)$ (\secref{first-order:lifting}) is also preserved by
% $H$. \roly{don't think we actually show this}

% The correctness of this interpretation is considered in the next section.

\subsection{Examples}

Let the signature
$\Sigma_{\mathsf{num}} = (\{\mathsf{num}\}, \{\mathrm{zero} : \tyUnit \to \mathsf{num}, \mathrm{add} :
\mathsf{num} \times \mathsf{num} \to \mathsf{num}\})$. This signature suffices to write the simple query
function in \exref{introduction-example}, where we interpret the $\mathrm{Label}$ type as the sum
$\tyUnit \tySum \tyUnit$ and the labels $\mathsf{a}$ and $\mathsf{b}$ as $\tmInl{\tmUnit}$ and
$\tmInr{\tmUnit}$. We consider several interpretations of $\Sigma_{\mathsf{num}}$ in $\Fam(\LatGal)$ and their
behaviour on the selection-and-sum query from \exref{introduction-example}. First, we note the type of the
reverse approximation map in this case. The type of the approximation maps depends on the input value. Our example input database was $\mathit{db} = [(\mathsf{a}, 0), (\mathsf{b}, 1), (\mathsf{a}, 1)]$, meaning that the type of the reverse approximation map for this database and any label $l$ is:
\begin{displaymath}
  \begin{array}{l}
    (\partial \mathrm{query})_r(l, \mathit{db}) : \\
    \quad \partial\sem{\mathsf{num}}(\mathrm{query}\,(l,\mathit{db})) \multimap \mathbb{1} \times (\mathbb{1} \times \partial\sem{\mathsf{num}}(0)) \times (\mathbb{1} \times \partial\sem{\mathsf{num}}(1)) \times (\mathbb{1} \times \partial\sem{\mathsf{num}}(1)) \times \mathbb{1}
  \end{array}
\end{displaymath}
in the category $\JoinSLat$, where $\partial\sem{\mathsf{num}}(x)$ is the lattice of approximations of the
number $x$ determined by our interpretation. In the codomain, the first four $\mathbb{1}$s correspond to the
positions of labels in the input, which we are not approximating, and the final $\mathbb{1}$ is the terminator
of the list. Note how, even if the $\partial\sem{\mathsf{num}}(x)$ does not depend on $x$ the type of the
output is still dependent on the shape of the input list: type dependency is used in a fundamental way in our
interpretation.
\newcommand{\TWO}{\mathbb{2}}

\begin{enumerate}
\item If we take $\sem{\mathsf{num}}_{\PrimTy} = \Disc(\mathbb{R})$, with $\sem{\mathrm{zero}}_{\PrimOp}$ and
  $\sem{\mathrm{add}}_{\PrimOp}$ the embeddings of the usual zero and addition functions via $\Disc$, then the
  resulting interpretation contains no approximation information. We have
  $\partial\sem{\mathsf{num}}(x) = \mathbb{1}$ so the type of $(\partial \mathrm{query})_r$ is trivial.
\item We take $\sem{\mathsf{num}}_{\PrimTy} = \Lift(\Disc(\mathbb{R}))$, using the lifting monad from
  \secref{models-of-total-gps:lifting}, with $\sem{\mathrm{zero}}_{\PrimOp}$ and
  $\sem{\mathrm{add}}_{\PrimOp}$ defined from the unlifted interpretations above and the monad structure. The
  type of the reverse map now becomes, where $\TWO = \{\top.\bot\}$:
  \begin{equation}\label{eqn:reverse-query-approx-1}
    (\partial \mathrm{query})_r(l, \mathit{db}) : \TWO \multimap \mathbb{1} \times (\mathbb{1} \times \TWO) \times (\mathbb{1} \times \TWO) \times (\mathbb{1} \times \TWO) \times \mathbb{1}
  \end{equation}
  where every position that corresponds to a number has been tagged with $\top$ for ``present'' and $\bot$ for
  ``not present''. This interpretation recovers the behaviour given in \exref{introduction-example}: running
  the reverse approximation map at the input ``$\mathsf{a}$'' at approximation $\top$ reveals that only the
  numbers in the rows tagged with ``$\mathsf{a}$'' in the input are used, and likewise for ``$\mathsf{b}$''.
\item Quantitative approximation information with non-trivial dependency can be obtained by using the interval
  approximation interpretation from \exref{intervals-and-maxima-elements} and \secref{interval-approx}. We let
  $\sem{\mathsf{num}}_{\PrimTy} = \mathbb{R}_{\mathit{intv}}$ and interpret addition using the morphism given
  in \secref{interval-approx}. Recall that $\mathrm{query}\,(\mathsf{a},\mathit{db}) = 1$, so in the reverse
  direction we must choose an interval containing $1$ to discover the largest (i.e. least in the order)
  intervals that will give rise to this output as \emph{independent} changes to the input. For example, if we
  pick $[\frac{9}{10},\frac{11}{10}]$ as the interval, then:
  \begin{displaymath}
    (\partial\mathrm{query})_r(\mathsf{a},\mathsf{db})([\frac{9}{10},\frac{11}{10}]) = \cdot, (\cdot,[-\frac{1}{10},\frac{1}{10}]),(\cdot,\bot),(\cdot,[\frac{9}{10},\frac{11}{10}]), \cdot
  \end{displaymath}
  Thus, to achieve an output within $[\frac{9}{10},\frac{11}{10}]$, either the first ``$\mathsf{a}$'' row
  could be in $[-\frac{1}{10},\frac{1}{10}]$ or the second one could be in $[\frac{9}{10},\frac{11}{10}]$, and
  the number in the ``$\mathsf{b}$'' row is not relevant.
\end{enumerate}

We have tested these examples on our Agda implementation. See the file \texttt{example.agda}.

\subsection{Systematic Insertion of Approximation via Moggi's CBN translation}
\label{sec:cbn-translation}

We can now carry out the systematic insertion of approximation points that we foreshadowed in
\secref{models-of-total-gps:lifting}, using \citet[\S 3.1]{notions-of-computation}'s monadic CBN
translation. We use the $T$ monad from \secref{tagging-monad} because it can be defined in terms of the
language constructs we already have. This requires that we have a signature $\Sigma$ that includes a
primitive type to be interpreted as the approximation object $\mathbb{A}$ and primitive operations to be
interpreted as the monoid operations on this object.

The monadic CBN translation is standard, and entirely determined by the translation on types, so we only
define the type translation $\cbn{-}$ here:
\begin{mathpar}
  \cbn{\rho} = \rho

  \cbn{\sigma \tySum \tau} = T(\cbn{\sigma}) \tySum T(\cbn{\tau})

  \cbn{\tyUnit} = \tyUnit

  \cbn{\sigma \tyProd \tau} = T(\cbn{\sigma}) \tyProd T(\cbn{\tau})

  \cbn{\sigma \tyFun \tau} = T(\cbn{\sigma}) \tyFun T(\cbn{\tau})

  \cbn{\tyList\;\tau} = \tyList\,(T(\cbn{\tau}))
\end{mathpar}
Thus, the CBN translation on types inserts a use of the monad $T(X) = \mathbb{A} \times X$ at the point ``just
underneath'' every type former. In this case, we are describing the type of data annotated at every
level. Note that our lists here are still ``strict'', an alternative approach would be to consider ``lazy''
lists that wrap the tail of every node in a $T$ as well.

\newcommand{\Element}{(\TWO \times \TWO)}

We illustrate the effect of the CBN translation on the query example from
\exref{introduction-example}. Applying $\cbn{-}$ to the type of $\mathrm{query}$ yields:
\begin{displaymath}
  T(\mathrm{Label}) \times T(\tyList(T(T(\mathrm{Label}) \tyProd T(\mathsf{num})))) \to T(\mathsf{num})
\end{displaymath}
Thus, every substructure of the input and output has been annotated with usage information. Under
interpretation in $\Fam(\MeetSLat \times \JoinSLat^\op)$, the type of the reverse approximation map as a
morphism in $\JoinSLat$ at the input $\mathit{db}$ is now (suppressing some ``$\times \mathbb{1}$'' for
readability):
\begin{displaymath}
  (\partial \mathrm{query})_r(l, \mathit{db}) : \TWO \multimap \TWO \times (\TWO \times \Element \times \Element \times \Element)
\end{displaymath}
Comparing to the type in (\ref{eqn:reverse-query-approx-1}), we now gain much more fine-grained information on
which parts of the input are used. When $l = \mathsf{a}$, we have
$(\partial \mathrm{query})_r(\mathsf{a},\mathit{db})(\top) = (\top, (\top, (\top, \top), (\top, \bot), (\top,
\top)))$, indicating that the execution of this query had examined everything except the number in the second
entry of the database. With the previous interpretation, we did not have confirmation that the labels in each
row were actually required.
