\subsection{Biproducts}
\label{sec:biproduct}

\begin{definition}[Zero object]
A \emph{zero} object is an object which is both terminal and initial.
\end{definition}

\begin{definition}[Biproduct]
Suppose $C$ a $\CMon$-enriched category. For any objects $A, B$, the \emph{biproduct} of $A$ and $B$ is an
object $A \biprod B$ of $C$ together with morphisms

\begin{center}
\begin{tikzcd}
   A \arrow[r, "\biinj_A", shift left] &
   A \biprod B \arrow[l, "\biproj_A", shift left] \arrow[r, "\biproj_B"', shift right] &
   B \arrow[l, "\biinj_B"', shift right]
\end{tikzcd}
\end{center}

\noindent satisfying

\begin{minipage}[t]{0.45\textwidth}
\begin{center}
\begin{salign*}
   \biproj_A \comp \biinj_A &= \id_A \\
   \biproj_B \comp \biinj_A &= \zero_{A,B}
\end{salign*}
\end{center}
\end{minipage}%
\begin{minipage}[t]{0.45\textwidth}
\begin{center}
\begin{salign*}
   \biproj_B \comp \biinj_B &= \id_B \\
   \biproj_A \comp \biinj_B &= \zero_{B,A}
\end{salign*}
\end{center}
\end{minipage}

\noindent and

\begin{salign*}
\biinj_A \comp \biproj_A + \biinj_B \comp \biproj_B &= \id_{A \biprod B}
\end{salign*}
\end{definition}

\noindent \todo{I think we can also define ``biproduct'' just using zero morphisms, and then derive rather
than assume $\CMon$-enrichment.}

We say a $\CMon$-enriched $C$ is \emph{semi-additive} if it also has finite biproducts; having products or
coproducts is sufficient.

\begin{proposition}
\label{prop:biproduct:from-product-or-coproduct}
Suppose $C$ is $\CMon$-enriched.
\begin{itemize}
\item If $C$ has biproducts $(A \biprod B, \biinj_A, \biinj_B, \biproj_A, \biproj_B)$ then $(A
\biprod B, \biproj_A, \biproj_B)$ is a product and $(A \biprod B, \biinj_A, \biinj_B)$ is a coproduct.
\item If $C$ has binary products $(A \times B, \proj_1, \proj_2)$ then:
\begin{enumerate}
\item $\prodM{0_{A,B}}{0_{A,C}} = 0_{A,B \times C}$
\item $\prodM{f_1 + g_1}{f_2 + g_2} = \prodM{f_1}{f_2} + \prodM{g_1}{g_2}$
\item $(A \times B, \biinj_A, \biinj_B, \proj_1, \proj_2)$ is a
biproduct with $\biinj_A: A \to A \times B = \prodM{\id_A}{0_{A,B}}$ and $\biinj_B: B \to A \times B =
\prodM{0_{A,B}}{\id_B}$.
\end{enumerate}
\item If $C$ has binary coproducts $(A + B, \inj_1, \inj_2)$ then
\begin{enumerate}
\item $\coprodM{0_{A,C}}{0_{B,C}} = 0_{A + B,C}$
\item $\coprodM{f_1 + g_1}{f_2 + g_2} = \coprodM{f_1}{f_2} + \coprodM{g_1}{g_2}$
\item $(A + B, \inj_1, \inj_2, \biproj_A, \biproj_B)$ is a
biproduct with $\biproj_A: A + B \to A = \coprodM{\id_A}{0_{B,A}}$ and $\biproj_B: A + B \to B =
\coprodM{0_{A,B}}{\id_B}$.
\end{enumerate}
\item If $C$ has a terminal (or initial) object then it is a zero object.
\end{itemize}
\end{proposition}

\begin{proposition}
\label{prop:biproduct:prod-coprod}
Suppose $C$ a category with biproducts. Then the following commutes for any morphisms $f_1, g_1, f_2,
g_2$ in $C$:

\begin{center}
\begin{tikzcd}[row sep=3em, column sep=3em]
   & A \arrow[d, "\biinj_A", shift left] \arrow[dr, "f_2"] \\
   A' \arrow[r, "\prodM{f_1}{g_1}"] \arrow[dr, "g_1"'] \arrow[ur, "f_1"] & A \biprod B \arrow[u, "\biproj_A", shift left] \arrow[d, "\biproj_B"', shift right] \arrow[r, pos=0.35, "\coprodM{f_2}{g_2}"] & B' \\
   & B \arrow[u, "\biinj_B"', shift right] \arrow[ur, "g_2"']
\end{tikzcd}
\end{center}
\end{proposition}
