\section{Matrices over a commutative semiring}
\label{sec:matrix}

Suppose $S = (S, +, 0, \cdot, 1)$ a commutative semiring. Write $\sum_i a_i$ for $a_1 + \cdots + a_n$. Two
derived properties of $\sum$ will be useful: \emph{distributivity} of $\cdot$ over $\sum$, $a \cdot \sum_i b_i
= \sum_i a \cdot b_i$ (and symmetrically on the right), which follows from the distributivity of $\cdot$ over
$+$, and the \emph{interchange law} $\sum_i \sum_j a_{ij} = \sum_j \sum_i a_{ij}$, which follows from
associativity and commutativity of $+$.

\subsection{Vectors, matrices, and standard basis}
\label{sec:matrix:vectors}

An \emph{$n$-length vector over $X$} is an element $v \in X^n$, with $i$-th component $v_i$. An \emph{$n
\times m$ matrix over $S$} is an $m$-length vector of $n$-length vectors over $S$ (its columns), with entry
$M_{ij}$ at row $i$ and column $j$; equivalently, an $n$-length vector of $m$-length vectors over $S$ (its
rows). The \emph{standard basis vector} $e_i \in S^n$ at index $i$ is the Kronecker $\delta$,
\begin{salign*}
   (e_i)_j =
      \begin{cases}
         1 & \text{if }i = j; \\
         0 & \text{otherwise}
      \end{cases}
\end{salign*}
\noindent The scalar \emph{dot product} of two same-length vectors is $u \cdot v = \sum_i u_i \cdot v_i$.
Viewing vectors $w_1, \ldots, w_m \in S^n$ as the columns of an $n \times m$ matrix $W$ (so $W_{ji} =
(w_i)_j$), the iterated sum $\sum_i w_i \in S^n$ is the row-sum of $W$, defined componentwise from scalar
sum:
\begin{salign*}
   \left(\sum_{i} w_i\right)_{\!j} = \sum_i W_{ji}.
\end{salign*}

\subsection{Basis decomposition}
\label{sec:matrix:basis-decomp}

Write $a \cdot w$ for the pointwise product of a scalar $a$ with a vector $w$, i.e.\ $(a \cdot w)_i = a
\cdot w_i$. Every $m$-length vector $v$ decomposes as an iterated sum of scaled basis vectors:
\begin{salign*}
   v = \sum_{i} v_i \cdot e_i \tag{basis decomposition}
\end{salign*}
\noindent This follows from the symmetry $(e_i)_j = (e_j)_i$ of the Kronecker $\delta$ together with the
following lemma.

\begin{lemma}[Sum-of-basis]
\label{lem:matrix:sum-of-basis}
$\sum_j (e_i)_j \cdot v_j = v_i$.
\end{lemma}

\subsection{The matrix category}
\label{sec:matrix:category}

The product $NM$ of an $n \times p$ matrix $N$ with a $p \times m$ matrix $M$ is the $n \times m$ matrix
\begin{salign*}
   (NM)_{ij} = \sum_k N_{ik} \cdot M_{kj}.
\end{salign*}
\noindent This product is associative (using distributivity and the interchange law), with the $m \times m$
Kronecker $\delta$ matrix $I_{ij} = (e_i)_j$ as left and right unit (by \lemref{matrix:sum-of-basis}).
Accordingly, the category $\Mat(S)$ has:
\begin{itemize}
\item as objects all natural numbers;
\item as morphisms $m \to n$ all $n \times m$ matrices over $S$.
\end{itemize}
Composition $N \comp M$ is matrix multiplication $NM$, with $I_m$ as the identity on $m$.

\subsubsection{Dagger structure}
\label{sec:matrix:transpose}

The \emph{transpose} $M^{\!\top} : n \to m$ of $M : m \to n$ has entries $(M^{\!\top})_{ji} = M_{ij}$. It is
involutive, $(M^{\!\top})^{\!\top} = M$, and interacts with composition as $(N \comp M)^{\!\top} = M^{\!\top}
\comp N^{\!\top}$ (both immediate from commutativity of $\cdot$), so $(-)^{\!\top}$ is a dagger on
$\Mat(S)$.

\subsubsection{$\CMon$-enrichment and biproducts}
\label{sec:matrix:cmon}

Pointwise addition $(M + N)_{ij} = M_{ij} + N_{ij}$ and the zero matrix $0$ equip each hom-set of $\Mat(S)$
with a commutative monoid; composition is bilinear (distributes over $+$ on both sides), so that $\Mat(S)$
is $\CMon$-enriched.

Moreover, the natural number $0$ is a zero object (both terminal and initial), and for natural numbers $m$
and $n$ the biproduct $m \biprod n$ can be constructed as the sum $m + n$, with pairings and copairings
realised as vertical stacking and horizontal concatenation of matrices: for $M : k \to m$ and $N : k \to n$
the pairing $\prodM{M}{N} : k \to m+n$ stacks $M$ above $N$, and for $M : m \to k$ and $N : n \to k$ the
copairing $\coprodM{M}{N} : m+n \to k$ concatenates $M$ and $N$ side by side. The projections and
injections are correspondingly copairings and pairings of identity and zero:
\begin{salign*}
   \biproj_1 &= \coprodM{I_m}{0} = \begin{bmatrix} I_m & 0 \end{bmatrix} : m+n \to m, &
   \biproj_2 &= \coprodM{0}{I_n} = \begin{bmatrix} 0 & I_n \end{bmatrix} : m+n \to n, \\
   \biinj_1 &= \prodM{I_m}{0} = \begin{bmatrix} I_m \\ 0 \end{bmatrix} : m \to m+n, &
   \biinj_2 &= \prodM{0}{I_n} = \begin{bmatrix} 0 \\ I_n \end{bmatrix} : n \to m+n,
\end{salign*}
\noindent where $I_m, I_n$ are identity matrices and the $0$s are zero matrices of compatible size; the
relationship $\biinj_i = \biproj_i^{\!\top}$ makes the injections the dagger images of the projections.
$\Mat(S)$ is therefore a semi-additive category (\secref{biproduct}).

The iterated sum $\sum_i w_i$ from \secref{matrix:vectors} fits into this picture by considering a vector $w
\in S^n$ as a morphism $w : 1 \to n$ in $\Mat(S)$. For $w_1, \ldots, w_m \in S^n$, the iterated sum $\sum_i
w_i$ is their $\CMon$ addition in $\Mat(S)(1, n)$. Equivalently, $\sum_i w_i = W \comp \diag_m$, where $W : m
\to n$ has the $w_i$ as columns and $\diag_m : 1 \to m$ is the diagonal into the biproduct $1 \biprod \cdots
\biprod 1$.

\subsection{$\Mat(S)$ in any biproduct category}
\label{sec:matrix:representation}

Suppose $\cat{C}$ is a $\CMon$-enriched category with finite biproducts, $X \in \cat{C}$ a chosen object, and
$S \cong \cat{C}(X, X)$ a semiring isomorphism between $S$ and the endomorphism semiring of $X$ under
composition. Then $\Mat(S)$ is $\CMon$-enriched equivalent to the full subcategory of $\cat{C}$ whose objects
are the biproduct powers $X^n = X \biprod \cdots \biprod X$. An $n \times m$ matrix $M$ over $S$ corresponds
to the morphism $X^m \to X^n$ assembled from the entries $M_{ij} \in S \cong \cat{C}(X, X)$, and conversely
every such morphism is determined by its $X^n$-indexed components. Commutativity of $\cdot$ in $S$ means
composition is commutative in $\cat{C}(X, X)$.

For example, with $\cat{C} = \SemiLat$ and $X = \Two$, the isomorphism $\Two \cong \SemiLat(\Two, \Two)$
identifies $\bot$ with the zero morphism and $\top$ with the identity; every join-preserving endomorphism of
$\Two$ is determined by its value at $\top$. This yields $\Mat(\Two)$ as $\CMon$-enriched equivalent to the
full subcategory of $\SemiLat$ on objects $\Two^n$.

\subsection{Additional structure on $S$}
\label{sec:matrix:additional-structure}

Further constructions become available when $S$ carries compatible order-theoretic structure.

\subsubsection{Join-semilattice}
\label{sec:matrix:additional-structure:join}

If $+$ is idempotent ($a + a = a$), then $(S, +, 0)$ is a join-semilattice with order $a \le b \iff a + b =
b$, and $\sum_i$ becomes iterated join (both on scalars and pointwise on vectors). In this setting the basis
decomposition of vectors (\secref{matrix:basis-decomp}) extends to \emph{linear} maps $f : S^m \to S^n$: those
that preserve $+$ (so $f(u + v) = f(u) + f(v)$ and $f(0) = 0$) and commute with scalar multiplication ($f(a
\cdot v) = a \cdot f(v)$), which is the usual notion of linearity extended to semirings. Every such $f$
satisfies
\begin{salign*}
   f(v) = \sum_i v_i \cdot f(e_i) \tag{basis decomposition of linear maps}
\end{salign*}
\noindent i.e.\ $f(v)$ is a ``join of atomic slices''. So $f$ is determined by the $m$ vectors $f(e_1),
\ldots, f(e_m) \in S^n$. These form the columns
of an $n \times m$ matrix $M_f$, the \emph{matrix representation} of $f$; conversely every $n \times m$
matrix is $M_f$ for a unique such $f$.

\subsubsection{Distributive lattice}
\label{sec:matrix:additional-structure:distrib}

If additionally $\cdot$ is also idempotent and $1 + a = 1$ for all $a$ (so $1$ is the additive top), then
$(S, +, 0, \cdot, 1, \le)$ is a bounded distributive lattice: $\cdot$ is the binary meet, $1$ is the top, $+$
is the binary join (from \secref{matrix:additional-structure:join}), and $0$ is the bottom. Distributivity
of $\cdot$ over $+$ is already a semiring axiom. To make the lattice perspective explicit, we use $\join =
+$, $\meet = \cdot$, $\bot = 0$ and $\top = 1$ in what follows; for iterated joins we write $\bigjoin_i a_i
= \sum_i a_i$ and $\bigmeet_i a_i$ for the iterated meet (= iterated $\sum$ in $\Mat(S^\op)$). Disjointness
$a \disj b \iff a \meet b \le \bot$ lifts pointwise to vectors.

\paragraph{Conjugate pair in $\LatConj$.} A linear $f : S^m \to S^n$ in $\Mat(S)$ has a transpose $f^{\!\top}
: S^n \to S^m$ acting as $f^{\!\top}(y)_j = \bigjoin_i M_{ij} \meet y_i$. The conjugacy condition
\[ y \disj f(x) \iff f^{\!\top}(y) \disj x \] holds for any matrix $M$ via distributivity, so $(f^{\!\top},
f)$ is a conjugate pair in $\LatConj$ on the appropriate lattices. Both directions preserve joins.

\paragraph{Opposite semiring and the adjoint candidate.} The opposite semiring $S^\op = (S, \meet, \top,
\join, \bot)$ is itself a bounded distributive lattice, with the roles of $\join$ and $\meet$ swapped.
Interpreting matrix multiplication in $\Mat(S^\op)$ on $M^{\!\top}$ gives the corresponding action $V_n \to V_m$:
\begin{salign*}
   M^{\!\top}_{S^\op}(y)_j \;=\; \bigmeet_i (M_{ij} \join y_i),
\end{salign*}
\noindent Distributivity ensures this map is meet-preserving:
\begin{salign*}
   M^{\!\top}_{S^\op}(y_1 \meet y_2)_j
   &= \bigmeet_i (M_{ij} \join (y_{1i} \meet y_{2i}))
   = \bigmeet_i \big((M_{ij} \join y_{1i}) \meet (M_{ij} \join y_{2i})\big) \\
   &= M^{\!\top}_{S^\op}(y_1)_j \meet M^{\!\top}_{S^\op}(y_2)_j.
\end{salign*}

\noindent So $M^{\!\top}_{S^\op}$ is a meet-semilattice homomorphism $V_n \to V_m$, and therefore right
adjoint to \emph{some} join-preserving map. Whether it is the right adjoint of $M$ depends on whether the
lattice supports the additional scalar structure (Heyting implication); see
\secref{matrix:additional-structure:boolean}. $M^{\!\top}_{S^\op}$ is perhaps best read as an adjoint
``candidate'': the automatically obtained meet-preserving dual of $M$ that becomes a genuine right adjoint
only if enough scalar structure is present.

\subsubsection{Boolean algebra and the Galois adjunction}
\label{sec:matrix:additional-structure:boolean}

If $S$ is furthermore a \emph{Boolean algebra} $B$, equipped with an operation $\neg$ satisfying $1 \le a +
\neg a$ (excluded middle) and $a \cdot \neg a \le 0$ (non-contradiction), then $\neg$ is involutive and
disjointness reflects the order. The conjugate pair $(f^{\!\top}, f)$ from
\secref{matrix:additional-structure:distrib} now lives inside the full subcategory of $\LatConj$ on Boolean
algebras, where the De~Morgan duality gives an equivalent presentation as a Galois adjunction in $\LatGal$:

\begin{salign*}
   f \adj f_*, \qquad f_*(x)_i = \bigmeet_j (M_{ij} \to x_j),
\end{salign*}
\noindent where $a \to b = \neg a \join b$ is Boolean implication. The right adjoint $f_*$ is
\emph{extensionally equal} to the opposite-semiring action $M^{\!\top}_{S^\op}$ applied to the
\emph{entrywise-complemented} transpose:
\begin{salign*}
   f_*(x) \;=\; (\neg M)^{\!\top}_{S^\op}(x)
   \qquad \text{i.e.\ }\;\; f_*(x)_i = \bigmeet_j (\neg M_{ij} \join x_j).
\end{salign*}

\noindent So the Boolean ingredient is precisely the entrywise complementation $\neg M$; everything else
(transposition, opposite-semiring action, meet-preservation) is already given by the distributive-lattice
construction. $\Mat(S)$-natively, then, Galois slicing factors as ``compute the $S^\op$-action of
$M^{\!\top}$'' plus ``apply $\neg$ to the matrix entries''. Without the Boolean structure, neither of these
two steps is available — the matrix-level analogue of the scalar-adjunction obstruction.

\subsubsection{$\Mat(\Two)$ and $\FinRel$}
\label{sec:matrix:two}

If $S$ is the Boolean algebra $\Two = (\{\bot, \top\}, \join, \bot, \meet, \top)$ with $+ := \join$, $\cdot :=
\meet$, $0 := \bot$ and $1 := \top$, then $\Mat(\Two)$ is equivalent to the category $\FinRel$ of finite sets
and binary relations. An $n \times m$ matrix $M : m \to n$ encodes the relation $R_M \subseteq [n] \times [m]$
via $i\, R_M\, j \iff M_{ij} = \top$, where $[n] = \{1, \ldots, n\}$. Matrix multiplication is relational
composition,
\begin{salign*}
   (N \comp M)_{ij} = \bigjoin_k N_{ik} \meet M_{kj},
\end{salign*}
\noindent which evaluates to $\top$ iff there is some intermediate $k$ with $j\, R_M\, k$ and $k\, R_N\, i$.
The dagger $M \mapsto M^{\!\top}$ is relational converse $R \mapsto R^{-1}$ on $\FinRel$, so $R_{M^{\!\top}} =
{R_M}^{-1}$. (More generally $\Mat(S)$ is equivalent to a category of finite sets and \emph{$S$-weighted}
relations, with matrix multiplication corresponding to weighted composition.)

\subsection{Higher-order model}
\label{sec:matrix:ho-model}

The matrix presentation lifts to a higher-order interpretation via the $\Fam$ construction
(\secref{fam}). $\Fam(\cat{C})$ is Cartesian closed only when $\cat{C}$ has set-indexed products, so the base
category for the slicing interpretation is generally not $\cat{C}$ itself but a target $\cat{D}$ with
sufficient products, into which $\cat{C}$ embeds via a finite-product-preserving functor $F : \cat{C} \to
\cat{D}$:

\begin{itemize}
\item For $\Mat(\Two)$, the target is $\JoinSLat$ itself, and $F$ sends $n \mapsto \Two^n$.

\item For $\LatConj$ (or its full subcategory on Boolean algebras), the target is $\JoinSLat \times
   \JoinSLat^{\op}$. $F$ forgets the lattice structure on objects; on morphisms it is the identity.

\item For $\LatGal$ (or its full subcategory on Boolean algebras), the target is $\MeetSLat \times
   \JoinSLat^{\op}$. $F$ forgets the lattice structure on objects; on morphisms it is the identity.
\end{itemize}

More generally, the construction applies to any $\Mat(S)$ embeddable into a biproduct category $\cat{C}$ as
in \secref{matrix:representation}, provided $\cat{C}$ in turn embeds into a category $\cat{D}$ with
set-indexed products.

\subsection{Connection to forward- and reverse-mode AD}
\label{sec:matrix:two:ad}

The conjugate and Galois slicing settings each give a forward and a backward map; together they populate the
cells of the following grid:

\begin{center}
\small
\begin{tabular}{lll}
 & \textbf{join-preserving} & \textbf{meet-preserving} \\
 fwd & $f^{\!\top}: \Two^m \to \Two^n$ \quad (Conjugate) & $f_*: \Two^m \to \Two^n$ \quad (Galois) \\
 bwd & $f: \Two^n \to \Two^m$ \quad (Galois) & $f^{\!\top}_*: \Two^n \to \Two^m$ \quad (Conjugate)
\end{tabular}
\end{center}

\noindent The diagonal pairs are adjunctions $f \dashv f_*$ and $f^{\!\top} \dashv f^{\!\top}_*$. The
intuition for the meet-preserving backward $f^{\!\top}_*$: where the join-preserving forward $f^{\!\top}$
propagates availability (``which inputs are present, hence which outputs are determined''), its right adjoint
asks a dual ``demand-relaxation'' question — \emph{``which inputs become unneeded if we no longer need these
outputs?''}.

For the AD comparison, we take the given morphism $f$ to be the backward direction in both slicing columns,
since it's join-preserving in both cases:

\begin{center}
\vspace{2mm}
\small
\begin{tabular}{llll}
 & \textbf{Automatic differentiation} & \textbf{Conjugate slicing} & \textbf{Galois slicing} \\
 fwd & $\pushf{\phi}_x: \RR^m \linearto \RR^n$ & $f^{\!\top}: \Two^m \to \Two^n$ & $f_*: \Two^m \to \Two^n$ \\
 category & $\FinVect_\RR$ & $\JoinSLat$ & $\MeetSLat$ \\
 via matrix mult in &$(\RR, +, \cdot)$ & $(\Two, \join, \meet)$ & $(\Two, \meet, \join)$ \\[2mm]
 bwd & $\pullf{\phi}_x: \RR^n \linearto \RR^m$ & $f: \Two^n \to \Two^m$ & $f: \Two^n \to \Two^m$ \\
 category & $\FinVect_\RR$ & $\JoinSLat$ & $\JoinSLat$ \\
 via matrix mult in &$(\RR, +, \cdot)$ & $(\Two, \join, \meet)$ & $(\Two, \join, \meet)$
\end{tabular}
\end{center}

\subsection{Example: rational intervals}
\label{sec:matrix:rational-intervals}

As an example, we will consider the 4 entries in the adjoint/conjugate grid on intervals over the rationals
$\QQ$. Let $\Intv(q) = \{[l, u] : l \leq q \leq u\}$ be the set of \emph{intervals} on $q \in $ ordered by
reverse inclusion (so a tighter interval is higher in the information order). $\Intv(q)$ is a distributive
upper-bounded lattice with top $[q, q]$, meet $[l_1, u_1] \meet [l_2, u_2] = [\max(l_1, l_2), \min(u_1, u_2)]$
and join $[l_1, u_1] \join [l_2, u_2] = [\min(l_1, l_2), \max(u_1, u_2)]$. The lifting $\IntvL(q) =
\Intv(q)_{\bot}$ which adjoins an explicit least element $\bot$ is a bounded distributive lattice, with $\bot$
as bottom.

For any $q_1, q_2 \in \QQ$, the addition $q_1 + q_2$ has four associated maps:

\begin{center}
\small
\begin{tabular}{ll}
\hline
\textbf{operation} & \textbf{behaviour} \\
\hline
$\add: \IntvL(q_1 + q_2) \to \IntvL(q_1) \times \IntvL(q_2)$ &
  $\add([l, u]) = $ \\
\emph{join-preserving backward map} & \quad $([l - q_2, u - q_2], [l - q_1, u - q_1])$ \\[1ex]
$\add^*: \IntvL(q_1) \times \IntvL(q_2) \to \IntvL(q_1 + q_2)$ &
  $\add^*([l_1, u_1], [l_2, u_2]) = $ \\
\emph{meet-preserving forward map} & \quad $[\min(q_2 + l_1, q_1 + l_2),\; \max(q_2 + u_1, q_1 + u_2)]$ \\[1ex]
$\addT: \IntvL(q_1) \times \IntvL(q_2) \to \IntvL(q_1 + q_2)$ &
  $\addT([l_1, u_1], [l_2, u_2]) = $ \\
\emph{join-preserving forward map} & \quad $[\max(q_2 + l_1, q_1 + l_2),\; \min(q_2 + u_1, q_1 + u_2)]$\\[1ex]
$\addT_*: \IntvL(q_1 + q_2) \to \IntvL(q_1) \times \IntvL(q_2)$ &
  $\addT_*([l, u]) = \add([l, u])$ \\
  \emph{meet-preserving backward map} & \\
\hline
\end{tabular}
\end{center}

\noindent The two input intervals each encode a constraint of the form ``this summand $q$ lies in the interval
$[l, u]$''; the forward maps $\add^*$ and $\addT$ shift each input by the \emph{other} summand $q'$ to obtain
candidate ranges for the result, and then combine them: by union for $\add^*$, or by intersection for $\addT$.

Note that $\addT_* = \add$, so in this case the four-map grid has only three distinct functions, with $\add$
being both left adjoint to $\add^*$ and right adjoint to $\addT$. This is possible because $\add$ is a lattice
homomorphism (preserves both joins and meets), and so can play both roles.

\paragraph{Comparison.} On overlapping but non-nested input intervals (e.g.\ $\IntvL$-shifted ranges $[1/2,
1]$ and $[4/5, 3/2]$):
\[
  \addT\colon\quad [\max(1/2, 4/5),\; \min(1, 3/2)] = [4/5, 1] \qquad \text{(intersection)}
\]
\[
  \add^*\colon\quad [\min(1/2, 4/5),\; \max(1, 3/2)] = [1/2, 3/2] \qquad \text{(union)}
\]

\paragraph{Connection to AD.} Forward-mode AD propagates tangents through addition by $dy = dx_1 + dx_2$.
Modulo idempotence of $+$ (replacing it by $\join$ in the information order), this corresponds to $\addT$:
each input contributes a constraint on the output, and the constraints are joined into the strongest
common refinement — exactly mirroring the role of $f^{\!\top}$ in the matrix-slicing comparison above.
$\add^*$ is the abstract-interpretation dual, mirroring $f_*$.
